DOpElib:\\ Differential Equations and Optimization Environment

### 2013-12-10 Examples/Documentation/foreword.tex
Update Doku. Habe Winnis-Superfunktionen fuer den Moment auskommentiert, da ich trotz tex-Update sonst nicht kompilieren kann

\twick{This is copied from the paper:
The main feature of \dope{} is to give a unified interface to high level 
algorithms such as 
time-stepping methods, nonlinear solvers and optimization routines. 
\dope{} is designed in such a way that the user only needs to write those parts
of the code that are problem dependent while all invariant 
parts of the algorithms
are reusable without any need for further coding.
In particular, the user is enabled to switch between various different 
algorithms without the need to rewrite the problem dependent code, 
though obviously he or she will
have to replace the algorithm object with an other one. 
This replacement can be done by replacing the appropriate object at only
one point in the code.
In addition to the finite element code provided by deal.II --which at present is the only FE-toolkit to which we provide an 
interface-- 
the presented library
\dope{} is user-focused by delivering
prefabricated tools which require adjustments by the user only for parts
connected to his specific problem. 
This is in contrast to deal.II which leaves the implementation of all high-level algorithms to the user.
The possible solution of 
optimization problems with PDE constraints is an innovative feature in \dope{}.
}

\twickremove{Please check if we can remove part and use the red part!
While deal.II leaves much of the work and many
decisions to the user, DOpE intends to be user-optimized by delivering
prefabricated tools which require from the user only adjustments connected
to his specific problem. The solution of optimal control problems with PDE
constraints is an innovation in the DOpE framework.
}

@@

\paragraph{Contributors \& Thanks:}
\added{We wish to express our gratitude to those people that have contributed to the library or
  have been providing open software to which we have provided interface.

@@

\todo[inline]{Wollen wir so einen Abschnitt? Fehlt da noch wer ? gcc ...?}

>>>>

The main feature of \dope{} is to give a unified interface to high level 
algorithms such as 
time-stepping methods, nonlinear solvers and optimization routines. 
\dope{} is designed in such a way that the user only needs to write those parts
of the code that are problem dependent while all invariant 
parts of the algorithms
are reusable without any need for further coding.
In particular, the user is enabled to switch between various different 
algorithms without the need to rewrite the problem dependent code, 
though obviously he or she will
have to replace the algorithm object with an other one. 
This replacement can be done by replacing the appropriate object at only
one point in the code.
In addition to the finite element code provided by deal.II --which at present is the only FE-toolkit to which we provide an 
interface-- 
the presented library
\dope{} is user-focused by delivering
prefabricated tools which require adjustments by the user only for parts
connected to his specific problem. 
This is in contrast to deal.II which leaves the implementation of all high-level algorithms to the user.
The possible solution of 
optimization problems with PDE constraints is an innovative feature in \dope{}.

@@

\paragraph{Contributors \& Thanks:}
{We wish to express our gratitude to those people that have contributed to the library or
  have been providing open software to which we have provided interface.

@@

### 2013-12-10 Examples/Documentation/introduction.tex
Update Doku. Habe Winnis-Superfunktionen fuer den Moment auskommentiert, da ich trotz tex-Update sonst nicht kompilieren kann

\chapter{Introduction}\label{chap:intro}
\section{How to get \dope{}}\label{sec:obtain}
\added{There are two ways to obtain a copy of \dope{}:\\
%
\begin{enumerate}[A)]
\item You can obtain a copy of \dope{} from the developers svn repository using\\
\texttt{svn co --username=[your username] \textbackslash \\ 
\hspace*{5mm}https://ganymed.iwr.uni-heidelberg.de/svn/dope}\\
(The backslash in the first line means that the line will be continued as one!)

Note that you need a valid username and password. If you have none you can contact the 
maintainer by sending an EMail to \url{dope@dopelib.net}.
%
\item You can download the sources as a tar-ball from the project website\\ 
\url{http://www.dopelib.net}.
\end{enumerate}
}

@@

\section{References}
\added{
If you like the library and use it for your own projects, please give credits by 
referencing the project \dope{}~\cite{dope} using the following bibtex entry:
}
\begin{lstlisting}
@MISC{dope,
  key   = {DOpElib},
  title = {The {D}ifferential {E}quation and 
               {O}ptimization {E}nvironment: \textsc{DOpElib}},
  url   = {http://www.dopelib.net},
  note  = {\texttt{http://www.dopelib.net}}
}
\end{lstlisting}

@@

\section{Installation}
\label{sec_installation}
I) Get a copy of \texttt{DOpElib}, see Section~\ref{sec:obtain} for details.
%
II) Install \texttt{deal.II} to your home directory, i.e., it should be located in \\
\texttt{\textasciitilde/deal.II}. \\[1mm]
If you would like to have another path you will either have to manually edit the files
\texttt{DOpEsrc/source/Makefile} and \texttt{Examples/Make.global\underline{ }options} and replace the 
line \\
\texttt{D = \$(HOME)/deal.II} with the appropriate path.\\[1mm]
Configure it properly and make the whole \texttt{deal.II} library.\\[3mm]
%
III) If you want to use some of the supported third party libraries install them and follow 
 the instructions in \texttt{ThirdPartyLibs/README}. There may be further information 
 in some \texttt{ThirdPartyLibs/*.INSTALLNOTES} that you may want to consider.\\[3mm]
%
IV) To build the \texttt{DOpElib} you have to change to the directory 
\texttt{DOpEsrc} where you can call\\ 
\texttt{make all}\\
to build the library. If you want to generate some documentation you may use
\texttt{make pdf-doc} or \texttt{make html-doc} which will require either 
latex or doxygen to be installed on your computer.\\[3mm]
%
V) To build the Examples change to the \texttt{Examples} directory and \texttt{make all}.
%
VI) Finally you may wish to test if everything worked. To do so you can 
change to the \texttt{Examples} directory and \texttt{make tests} which will give you a 
list of all the examples and whether they behave as expected by the library, see also 
Chapter~\ref{chap:test}.

>>>>

\chapter{Introduction}\label{chap:intro}
\section{How to get \dope{}}\label{sec:obtain}
There are two ways to obtain a copy of \dope{}:\\

\begin{enumerate}[A)]
\item You can obtain a copy of \dope{} from the developers svn repository using\\
\texttt{svn co --username=[your username] \textbackslash \\ 
\hspace*{5mm}https://ganymed.iwr.uni-heidelberg.de/svn/dope}\\
(The backslash in the first line means that the line will be continued as one!)

Note that you need a valid username and password. If you have none you can contact the 
maintainer by sending an EMail to \url{dope@dopelib.net}.
%
\item You can download the sources as a tar-ball from the project website\\ 
\url{http://www.dopelib.net}.
\end{enumerate}

@@

\section{References}
If you like the library and use it for your own projects, please give credits by 
referencing the project \dope{}~\cite{dope} using the following bibtex entry:

\begin{lstlisting}
@MISC{dope,
  key   = {DOpElib},
  title = {The {D}ifferential {E}quation and 
               {O}ptimization {E}nvironment: \textsc{DOpElib}},
  url   = {http://www.dopelib.net},
  note  = {\texttt{http://www.dopelib.net}}
}
\end{lstlisting}

@@

\section{Installation}
\label{sec_installation}
I) Get a copy of \texttt{DOpElib}, see Section~\ref{sec:obtain} for details.\\[1mm]
%
II) Install \texttt{deal.II} to your home directory, i.e., it should be located in \\
\texttt{\textasciitilde/deal.II}. \\[1mm]
If you would like to have another path you will either have to manually edit the files
\texttt{DOpEsrc/source/Makefile} and \texttt{Examples/Make.global\underline{ }options} and replace the 
line \\
\texttt{D = \$(HOME)/deal.II} with the appropriate path.\\[1mm]
Configure it properly and make the whole \texttt{deal.II} library.\\[3mm]
%
III) If you want to use some of the supported third party libraries install them and follow 
 the instructions in \texttt{ThirdPartyLibs/README}. There may be further information 
 in some \texttt{ThirdPartyLibs/*.INSTALLNOTES} that you may want to consider.\\[3mm]
%
IV) To build the \texttt{DOpElib} you have to change to the directory 
\texttt{DOpEsrc} where you can call\\ 
\texttt{make all}\\
to build the library. If you want to generate some documentation you may use
\texttt{make pdf-doc} or \texttt{make html-doc} which will require either 
latex or doxygen to be installed on your computer.\\[3mm]
%
V) To build the Examples change to the \texttt{Examples} directory and \texttt{make all}.
%
VI) Finally you may wish to test if everything worked. To do so you can 
change to the \texttt{Examples} directory and \texttt{make tests} which will give you a 
list of all the examples and whether they behave as expected by the library, see also 
Chapter~\ref{chap:test}.

### 2013-12-10 Examples/OPT/InstatPDE/Example1/content.tex
Update Doku. Habe Winnis-Superfunktionen fuer den Moment auskommentiert, da ich trotz tex-Update sonst nicht kompilieren kann

\subsubsection{Program description}
The following new things differ from the PDE-Examples and the optimization examples with stationary 
PDEs:\\[2mm]
\twick{
First, we need to introduce a reasonable dual time-stepping scheme in the main
file since we have to
compute 
the adjoint equation backward in time. 
Next,} in the file \texttt{main.cc} the SpaceTimeHandler now gets an additional argument.
In this case \texttt{DOpEtypes::initial} which specifies that the control is entering in the 
initial value.\\[2mm]
In the file \texttt{localpde.h} we now have to specify the Methods \texttt{Init\underline{ }ElementRhs}
and \texttt{Init\underline{ }ElementRhs\underline{ }Q}. They need to be adapted, since usually the InitialValue for the 
PDE is auto generated from a deal.ii function in the ProblemContainer. This Value is set in the 
\texttt{Init\underline{ }ElementRhs} hence we need to change this function to use the control instead. 
Correspondingly we need to implement the first derivative of this with respect to the control. 
--We don't need the second derivative since it is zero anyways.--\\[2mm]

>>>>

\subsubsection{Program description}
The following new things differ from the PDE-Examples and the optimization examples with stationary 
PDEs:\\[2mm]
First, we need to introduce a reasonable dual time-stepping scheme in the main
file since we have to
compute 
the adjoint equation backward in time. 
Next, in the file \texttt{main.cc} the SpaceTimeHandler now gets an additional argument.
In this case \texttt{DOpEtypes::initial} which specifies that the control is entering in the 
initial value.\\[2mm]
In the file \texttt{localpde.h} we now have to specify the Methods \texttt{Init\underline{ }ElementRhs}
and \texttt{Init\underline{ }ElementRhs\underline{ }Q}. They need to be adapted, since usually the InitialValue for the 
PDE is auto generated from a deal.ii function in the ProblemContainer. This Value is set in the 
\texttt{Init\underline{ }ElementRhs} hence we need to change this function to use the control instead. 
Correspondingly we need to implement the first derivative of this with respect to the control. 
--We don't need the second derivative since it is zero anyways.--\\[2mm]

### 2013-12-10 Examples/OPT/StatPDE/Example1/content.tex
Update Doku. Habe Winnis-Superfunktionen fuer den Moment auskommentiert, da ich trotz tex-Update sonst nicht kompilieren kann

\subsubsection{General problem description}
\twick{Dieses Beispiel ist nun extremst ausfuehrlich hinsichtlich der
localpde. Da dies ein Manual ist, wuensche ich mir, dass wir dies (respective
moeglicher Rechtschreibfehler und typos) auch so belassen. Mir hat das damals 
bei FSI Opt sehr geholfen.}
This example solves a distributed minimization problem
\twick{and shows how to estimate the error in the cost functional for stationary optimization
problems. The problem reads:}
\begin{gather*}
\min J(q,u) = \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2\\
\text{s.t.} (\nabla u,\nabla \phi) = (q+f,\phi)\;\;\forall\,\phi \in H^1_0(\Omega)
\end{gather*}
on the domain $\Omega = [0,1]^2$, and the data is chosen as follows:
\begin{gather*}
 f = \left(20\pi^2  \sin(4 \pi x) - \frac{1}{\alpha}  \sin(\pi x)\right) \sin(2 \pi y)\\
 u^d = \left( 5 \pi^2 \sin(\pi x) + \sin(4 \pi x)\right)  \sin(2\pi y)
\end{gather*}
and $\alpha = 10^{-3}$.
Hence its solution is given by:
\begin{gather*}
 \overline{q} = \frac{1}{\alpha} \sin(\pi x) \sin(2 \pi y)\\
 \overline{u} = \sin(4 \pi x) \sin(2 \pi y).
\end{gather*}
Thus the exact optimal value of the cost functional can be calculated as 
\begin{gather*}
 J^* = J(\overline{q},\overline{u}) = \frac{1}{8}\Bigl(25\pi^4 + \frac{1}{\alpha}\Bigr).
\end{gather*}

@@

\added{
\subsubsection{Background information and program description}
In the following, we describe all extensions to the previous problems 
relevant to solving PDE-based optimization with \dope{}. So far, 
we had only to implement the \texttt{ElementEquation} and the corresponding
matrix \texttt{ElementMatrix}. Now, based on the idea of the reduced cost
functional, we have to compute certain additional equations representing the 
adjoint, tangent, and adjoint hessian equations
\texttt{ElementEquation\underline{ }U, ElementEquation\underline{ }UT,
ElementEquation\underline{ }UTT}
for the state equation and in the same terms arising from the functional
itself. Let us shed some light into all equations by giving some background
information and overview first.

@@

Moreover, we would like to work in the Hilbert space $Q$. However, the derivative 
$j'(q) \in H^*$ only. Hence, we need to calculate the Riesz representation 
for the gradient $\nabla j(q) \in H$ using the definition:
\[
 (\nabla j(q),\delta q)_Q = j'(q)(\delta q) \quad \forall\,\delta q\in Q.
\]
In the given example, $Q = L^2(\Omega)$ and hence the scalar product will be the
standard $L^2$-inner product. Simillarly, we can define the Hessian operator 
$H(q) \in mathcal L(Q,Q)$ by defining
\[
 (H(q)\tau q,\delta q)_Q = j''(q)(\tau q,\delta q) \quad \forall\,\delta q,\tau q\in Q.
\]
}
\todo[inline]{Brauchen wir den n\"achsten Block wirklich? Das steht 1 Seite dar\"uber schon, 
die Notation passt aber nicht immer, z.B. $J''(u)$, fehlende Argumente und Funktionen (z.N. Dual Hessian Equation)
. 
Ansonsten müssen wir den Block nochmal durchgehen um die Sachen zu beheben.}

Finally, we recall the representations of the previous work. First, we recall 
in terms of the Lagrangian:

\textbf{Adjoint:} Find $z\in X$ such that
\begin{equation*}
{\cal L}_u' (q,u,z)(\phi) = 0 \quad \forall\phi\in X. 
\end{equation*}
\textbf{Tangent (linearized state):} Find $\delta u\in X$ such that
\begin{equation*}
{\cal L}_{qz}'' (q,u,z)(\delta q, \phi) +
{\cal L}_{uz}'' (q,u,z)(\delta u, \phi) = 0 \quad \forall\phi\in X. 
\end{equation*}

\textbf{Additional adjoint:} Find $\delta z\in X$ such that
\begin{equation*}
{\cal L}_{qu}'' (q,u,z)(\delta q, \phi) +
{\cal L}_{uu}'' (q,u,z)(\delta u, \phi) +
{\cal L}_{zu}'' (q,u,z)(\delta z, \phi) = 0 \quad \forall\phi\in X. 
\end{equation*}

Especially, the equations for stationary optimization problems are
given as follows:

\textbf{Adjoint:} Find $z\in X$ such that
\begin{equation*}
a_u'(q,u)(\phi,z) = J'(u)(\phi) \quad \forall\phi\in X. 
\end{equation*}
\textbf{Tangent:} Find $\delta u\in X$ such that
\begin{equation*}
a_u'(q,u)(\delta u,\phi) = -a_q'(q,u)(\delta q,\phi)  \quad \forall\phi\in X. 
\end{equation*}
\textbf{Additional adjoint:} Find $\delta z\in X$ such that
\begin{equation*}
a_u'(q,u)(\phi,\delta z) = a_{uu}''(q,u)(\delta u,\phi,z)
-a_{qu}''(q,u)(\delta q,\phi,z) + J''(u)(\delta,\phi) \quad \forall\phi\in X. 
\end{equation*}

Using the Lagrangian, the previous expressions read

\textbf{Lagrangian}
\begin{equation*}
{\cal L}(q,u,z) =J(q,u) - a(q,u)(z)
\end{equation*}

\textbf{State Equation}
\begin{align*}
a(q,u)(\phi) = 0 
\end{align*}

\textbf{Dual Equation}

\begin{align*}
{\cal L}_{u}' (q,u,z)(\phi ) &= 0 \\
\Leftrightarrow a_u' (q,u)(\phi,z) &= J_u' (q,u)(\phi)
\end{align*}

\textbf{Tangent Equation}

\begin{align*}
{\cal L}_{qz}'' (q,u,z)(\delta q, \phi) 
+ {\cal L}_{uz}'' (q,u,z)(\delta u, \phi) &= 0 \\
\Leftrightarrow a_u' (q,u)(\delta u,\phi) + a_q' (q,u)(\delta q,\phi) &= 0
\end{align*}

\textbf{Dual Hessian Equation}

\begin{align*}
{\cal L}_{qu}'' (q,u,z)(\delta q, \phi) 
&+ {\cal L}_{uu}'' (q,u,z)(\delta u, \phi) \\
&+ {\cal L}_{zu}'' (q,u,z)(\delta z, \phi) = 0 \\
\Leftrightarrow a_u' (q,u)(\phi, \delta z) 
&+ a_{uu}' (q,u)(\delta u,\phi, z) \\
&+ a_{qu}' (q,u)(\delta q,\phi, z)  = J_{uu}''(q,u)(\cdot ,\cdot )
\end{align*}

\textbf{Newton left hand side: first derivative}

\begin{align*}
j'(q)(\delta q) &= {\cal L}_{q}' (q,u,z)(\delta q) \quad\text{(Gradient equation)} \\
\Leftrightarrow j'(q)(\delta q) &= -a_q' (q,u)(\delta q,z) + J_{q}'(q,u)(\delta q)
\end{align*}

\textbf{Newton right hand side: second derivative}

\begin{align*}
j''(q)(\delta q, \delta r) &= {\cal L}_{qq}'' (q,u,z)(\delta q, \delta r) + 
{\cal L}_{uq}'' (q,u,z)(\delta u, \delta r) \\
&\quad +  {\cal L}_{zq}'' (q,u,z)(\delta z, \delta r) \\
\Leftrightarrow  j''(q)(\delta q, \delta r) &= 
J_{qq}''(q,u)(\delta q, \delta r) -a_{qq}'' (q,u)(\delta q,\delta r, z) \\
&\quad - a_{uq}'' (q,u)(\delta u,\delta r, z) -a_{q}' (q,u)(\delta r,\delta z)  
\end{align*}
\todo[inline]{end of delete comment}

\subsubsection{Implementation in \dope{}}
\added{
From the previous details, and the definition of the Lagrangian and its 
derivatives it is clear, that the user has to provide the respective derivatives. 
Since the Lagrangian consists of the PDE and the cost functional it 
is sufficient to provide the respective derivatives, while \texttt{DOpElib} 
will assemble them as required.
Test functions for vector valued terms will be denoted by an index $i$ while 
matrix valued terms are indexed in $i$ and $j$. Test functions in the control 
space $Q$ are denoted as $\phi^q_i$ while those in the state space $V$ are denoted 
as $\phi_i$.

@@

\subsubsection{Back to the specific equations in this example}
We have
\todo[inline]{Wollen wir die alle? Die Terme $a'_u$ sind ja alle Gleich bis auf das 
jeweilige argument...}
\begin{align*}
a(q,u)(\phi) &= (\nabla u, \nabla\phi) - (q + f, \phi),\\
a_u'(q,u)(\phi, z) &= (\nabla\phi, \nabla z),\\
a_u'(q,u)(\delta u, \phi) &= (\nabla\delta u, \nabla\phi),\\
a_u'(q,u)(\phi, \delta z) &= (\nabla\phi, \nabla\delta z),\\
a_q'(q,u)(\delta q, \phi) &= -(z,\psi^q),\\
a_q'(q,u)(\delta q, \phi) &= (\delta q,\phi),\\
a_{q}'(q,u)(\delta q,\delta z) &= - (\delta z,\psi^q).
\end{align*}
%
For the cost functional, we have the following terms:
\begin{align*}
J(q,u) &= \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2,\\
J_u'(q,u)(\phi) = (u-u^d, \phi),\\
J_q'(q,u)(\phi) = \alpha (q,\psi^q),\\
J_{uu}''(q,u)(\psi, \phi) = (\delta u, \phi).
\end{align*}
All other terms, specifically mixed terms with $QU$ etc. are zero in this
example.

>>>>

\subsubsection{General problem description}
This example solves a distributed minimization problem
and shows how to estimate the error in the cost functional for stationary optimization
problems. The problem reads:
\begin{gather*}
\min J(q,u) = \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2\\
\text{s.t.} (\nabla u,\nabla \phi) = (q+f,\phi)\;\;\forall\,\phi \in H^1_0(\Omega)
\end{gather*}
on the domain $\Omega = [0,1]^2$, and the data is chosen as follows:
\begin{gather*}
 f = \left(20\pi^2  \sin(4 \pi x) - \frac{1}{\alpha}  \sin(\pi x)\right) \sin(2 \pi y)\\
 u^d = \left( 5 \pi^2 \sin(\pi x) + \sin(4 \pi x)\right)  \sin(2\pi y)
\end{gather*}
and $\alpha = 10^{-3}$.
Hence its solution is given by:
\begin{gather*}
 \overline{q} = \frac{1}{\alpha} \sin(\pi x) \sin(2 \pi y)\\
 \overline{u} = \sin(4 \pi x) \sin(2 \pi y).
\end{gather*}
Thus the exact optimal value of the cost functional can be calculated as 
\begin{gather*}
 J^* = J(\overline{q},\overline{u}) = \frac{1}{8}\Bigl(25\pi^4 + \frac{1}{\alpha}\Bigr).
\end{gather*}

@@

{
\subsubsection{Background information and program description}
In the following, we describe all extensions to the previous problems 
relevant to solving PDE-based optimization with \dope{}. So far, 
we had only to implement the \texttt{ElementEquation} and the corresponding
matrix \texttt{ElementMatrix}. Now, based on the idea of the reduced cost
functional, we have to compute certain additional equations representing the 
adjoint, tangent, and adjoint hessian equations
\texttt{ElementEquation\underline{ }U, ElementEquation\underline{ }UT,
ElementEquation\underline{ }UTT}
for the state equation and in the same terms arising from the functional
itself. Let us shed some light into all equations by giving some background
information and overview first.

@@

Moreover, we would like to work in the Hilbert space $Q$. However, the derivative 
$j'(q) \in H^*$ only. Hence, we need to calculate the Riesz representation 
for the gradient $\nabla j(q) \in H$ using the definition:
\[
 (\nabla j(q),\delta q)_Q = j'(q)(\delta q) \quad \forall\,\delta q\in Q.
\]
In the given example, $Q = L^2(\Omega)$ and hence the scalar product will be the
standard $L^2$-inner product. Simillarly, we can define the Hessian operator 
$H(q) \in mathcal L(Q,Q)$ by defining
\[
 (H(q)\tau q,\delta q)_Q = j''(q)(\tau q,\delta q) \quad \forall\,\delta q,\tau q\in Q.
\]
}

\subsubsection{Implementation in \dope{}}
{
From the previous details, and the definition of the Lagrangian and its 
derivatives it is clear, that the user has to provide the respective derivatives. 
Since the Lagrangian consists of the PDE and the cost functional it 
is sufficient to provide the respective derivatives, while \texttt{DOpElib} 
will assemble them as required.
Test functions for vector valued terms will be denoted by an index $i$ while 
matrix valued terms are indexed in $i$ and $j$. Test functions in the control 
space $Q$ are denoted as $\phi^q_i$ while those in the state space $V$ are denoted 
as $\phi_i$.

@@

\subsubsection{Back to the specific equations in this example}
We have
\begin{align*}
a(q,u)(\phi) &= (\nabla u, \nabla\phi) - (q + f, \phi),\\
a_u'(q,u)(\phi, z) &= (\nabla\phi, \nabla z),\\
a_u'(q,u)(\delta u, \phi) &= (\nabla\delta u, \nabla\phi),\\
a_u'(q,u)(\phi, \delta z) &= (\nabla\phi, \nabla\delta z),\\
a_q'(q,u)(\delta q, \phi) &= -(z,\psi^q),\\
a_q'(q,u)(\delta q, \phi) &= (\delta q,\phi),\\
a_{q}'(q,u)(\delta q,\delta z) &= - (\delta z,\psi^q).
\end{align*}
%
For the cost functional, we have the following terms:
\begin{align*}
J(q,u) &= \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2,\\
J_u'(q,u)(\phi) = (u-u^d, \phi),\\
J_q'(q,u)(\phi) = \alpha (q,\psi^q),\\
J_{uu}''(q,u)(\psi, \phi) = (\delta u, \phi).
\end{align*}
All other terms, specifically mixed terms with $QU$ etc. are zero in this
example.

### 2013-12-10 Examples/OPT/StatPDE/Example2/content.tex
Update Doku. Habe Winnis-Superfunktionen fuer den Moment auskommentiert, da ich trotz tex-Update sonst nicht kompilieren kann

\subsubsection{General problem description}
This example solves the following pointwise minimization problem
\begin{align*}
\min_{(q,u)\in \mathbb R^3 \times H_0^1(\Omega; \mathbb R^2)} J(q,u) &= \frac{1}{2} \sum_{i=0}^2\abs{(u-\overline u)(x_i)}^2 + \frac{\alpha}{2}\|q\|^2&&\\
\text{s.t. } (\nabla u,\nabla \phi) &= (f(q),\phi) \quad\forall\,\phi \in H^1_0(\Omega; \mathbb R^2)
\end{align*}
on the domain $\Omega = [0,1]^2$, with zero Dirichlet boundary conditions and
\begin{itemize}
\item the observation points
\begin{align*}
x_0 = (0.5, 0.5), \quad x_1 = (0.5, 0.25),\quad x_2 = (0.25, 0.25),
\end{align*}
\item the regularization parameter $\alpha = 0$, 
\item the right hand side
\begin{align*}
 f(q) &= q_0 \left(\begin{matrix}2\pi^2  \sin( \pi x) \sin(\pi y)\\0 \end{matrix}\right)\\
      &+ q_1 \left(\begin{matrix}5\pi^2  \sin( \pi x) \sin(2\pi y)\\0 \end{matrix}\right)\\
      &+ q_2 \left(\begin{matrix}0 \\8\pi^2  \sin(2\pi x) \sin(2\pi y)\end{matrix}\right)\\
\end{align*}
\item and the exact solution given by 
\begin{align*}
 \overline{q} &= \bigl(1;0.5;1\bigr)\\
 \overline{u}& = \left(\begin{matrix} \sin( \pi x)( \sin(\pi y)+0.5\sin(2\pi y))\\\sin(2\pi x) \sin(2\pi y) \end{matrix}\right).
\end{align*}
\end{itemize}
\subsubsection{Program description}
\twick{
In contrast to the first example, the control is now a discrete quantity (for
the three observation points) where we use the \texttt{FE\underline{}Nothing}
element to assign the three controls. Here, the number of components equals
the number of controls. 
In addition, }notice that the cost functional is of mixed type (from our computational point of view), i.e. the first part is a pointfunctional whereas the regularization part requires the evaluation of a domain integral. To handle this, we need a special integrator as well as a special newtonsolver. Additionally, our \texttt{LocalFunctional} returns as his type:
\begin{verbatim}
      string
      GetType() const
      {
        return "point domain";
      }
\end{verbatim}
This brings along that we have not only to implement 
\begin{itemize} 
\item \texttt{ElementValue},
\item \texttt{ElementValue\_U},
\item \texttt{ElementValue\_Q}, 
\item \texttt{ElementValue\_UU},
\item \texttt{ElementValue\_UQ},
\item \texttt{ElementValue\_QU},
\item \texttt{ElementValue\_QQ}
\end{itemize}
 from \texttt{FunctionalInterface}, but also all the aforementioned methods with a preceding \text{Point} (\texttt{PointValue} etc.).
%FE_Nothing, handle of the control erlaeutern.

>>>>

\subsubsection{General problem description}
This example solves the following pointwise minimization problem
\begin{align*}
\min_{(q,u)\in \mathbb R^3 \times H_0^1(\Omega; \mathbb R^2)} J(q,u) &= \frac{1}{2} \sum_{i=0}^2\abs{(u-\overline u)(x_i)}^2 + \frac{\alpha}{2}\|q\|^2&&\\
\text{s.t. } (\nabla u,\nabla \phi) &= (f(q),\phi) \quad\forall\,\phi \in H^1_0(\Omega; \mathbb R^2)
\end{align*}
on the domain $\Omega = [0,1]^2$, with zero Dirichlet boundary conditions and
\begin{itemize}
\item the observation points
\begin{align*}
x_0 = (0.5, 0.5), \quad x_1 = (0.5, 0.25),\quad x_2 = (0.25, 0.25),
\end{align*}
\item the regularization parameter $\alpha = 0$, 
\item the right hand side
\begin{align*}
 f(q) &= q_0 \left(\begin{matrix}2\pi^2  \sin( \pi x) \sin(\pi y)\\0 \end{matrix}\right)\\
      &+ q_1 \left(\begin{matrix}5\pi^2  \sin( \pi x) \sin(2\pi y)\\0 \end{matrix}\right)\\
      &+ q_2 \left(\begin{matrix}0 \\8\pi^2  \sin(2\pi x) \sin(2\pi y)\end{matrix}\right)\\
\end{align*}
\item and the exact solution given by 
\begin{align*}
 \overline{q} &= \bigl(1;0.5;1\bigr)\\
 \overline{u}& = \left(\begin{matrix} \sin( \pi x)( \sin(\pi y)+0.5\sin(2\pi y))\\\sin(2\pi x) \sin(2\pi y) \end{matrix}\right).
\end{align*}
\end{itemize}
\subsubsection{Program description}
{
In contrast to the first example, the control is now a discrete quantity (for
the three observation points) where we use the \texttt{FE\underline{}Nothing}
element to assign the three controls. Here, the number of components equals
the number of controls. 
In addition, }notice that the cost functional is of mixed type (from our computational point of view), i.e. the first part is a pointfunctional whereas the regularization part requires the evaluation of a domain integral. To handle this, we need a special integrator as well as a special newtonsolver. Additionally, our \texttt{LocalFunctional} returns as his type:
\begin{verbatim}
      string
      GetType() const
      {
        return "point domain";
      }
\end{verbatim}
This brings along that we have not only to implement 
\begin{itemize} 
\item \texttt{ElementValue},
\item \texttt{ElementValue\_U},
\item \texttt{ElementValue\_Q}, 
\item \texttt{ElementValue\_UU},
\item \texttt{ElementValue\_UQ},
\item \texttt{ElementValue\_QU},
\item \texttt{ElementValue\_QQ}
\end{itemize}
 from \texttt{FunctionalInterface}, but also all the aforementioned methods with a preceding \text{Point} (\texttt{PointValue} etc.).
%FE_Nothing, handle of the control erlaeutern.

### 2013-12-10 Examples/OPT/StatPDE/Example6/content.tex
Finished overhaul of OPT StatPDE examples. 7 and 8 are untested due to missing snopt
Update Doku. Habe Winnis-Superfunktionen fuer den Moment auskommentiert, da ich trotz tex-Update sonst nicht kompilieren kann

This example implements the minimum compliance problem for the thickness optimization 
of an MBB-Beam. Using the MMA-Method of K. Svanberg together with an augmented Lagrangian 
approach for the subproblems following M. Stingl.

>>>>

\subsubsection{General problem description}
This example solves the distributed minimization problem
\begin{gather*}
\min J(q,u) = \frac{1}{2} \|u-u^d\|^2 + \frac{\alpha}{2}\|q\|^2\\
\text{s.t.} (\nabla u,\nabla \phi) = (q+f,\phi)\;\;\forall\,\phi \in H^1_0(\Omega)\\
\text{s.t.} -500 \le q \le 500\;\text{a.e. in }\Omega
\end{gather*}
on the domain $\Omega = [0,1]^2$, and the data is chosen as follows:
\begin{gather*}
 f = \left(20\pi^2  \sin(4 \pi x) - \frac{1}{\alpha}  \sin(\pi x)\right) \sin(2 \pi y)\\
 u^d = \left( 5 \pi^2 \sin(\pi x) + \sin(4 \pi x)\right)  \sin(2\pi y)
\end{gather*}
and $\alpha = 10^{-3}$.

In addition the following functionals are evaluated:
\begin{gather*}
  \text{MidPoint: } u(0.5 ; 0.5)\\[2mm]
  \text{MeanValue: }\int_\Omega u
\end{gather*}

\subsubsection{Program description}
{
The Problem is similar to that of {\tt OPT/StatPDE/Example1} except for the box control constraints.
The implementation of these constraints is taken care of in the main file
where we add a constrained description \texttt{lcc}.
}

\textcolor{blue}{Winni: Aus den numerischen 
Ergebnissen lese ich ab, das die Kontrolle zwischen 0 und 1 liegt.
\[
0 \leq q \leq 1.
\]
 Aber ich habe keine Ahnung was die Assoziierung zu 
\texttt{
lcc[0][0] = 1; //each component is constrained individualy
  lcc[0][1] = 2; // each two constraints (lower and upper bound)
}
ist. Auch die zugehoerigen Kommentare im main.cc file bringen mich nicht
  wirklich
weiter. Wieso each component? Das ist nur eine Kontrolle?! Oder meint
  component die state-equation component? 
Also konkret: ich kann nicht sagen, wofuer bei \texttt{lcc[0][0] = 1}  die 
einzelnen Eintraege stehen.
}

\subsubsection{External optimization solver}
The problem is solved using the optimization library IPOPT that you can obtain for free. To use it 
a correct link to the ipopt library needs to be created in  {\tt DOpE/ThirdPartyLibs} by the name 
{\tt ipopt}, i.e., you should have the file
{\tt DOpE/ThirdPartyLibs/ipopt} pointing to the ipopt directory. If you have not done this you can compile the
example but when running the example you will only get an error message like\\
{\tt Warning: During execution of `Reduced\underline{ }IpoptAlgorithm::Solve`\\ the following Problem occurred!\\
To use this algorithm you need to have IPOPT installed! \\To use this set the WITH\underline{ }IPOPT CompilerFlag.}

Alternatively the commercial optimization library SNOPT can be used in this example.
In order to use this library you need to install SNOPT on your computer and then generate a symlink to 
the snopt directory (where you have the libs and the header files) 
in the {\tt DOpE/ThirdPartyLibs} directory named {\tt snopt}, i.e., you should have the file
{\tt DOpE/ThirdPartyLibs/snopt} pointing to the snopt directory. If you have not done this you can compile the
example but when running the example you will only get an error message like\\
{\tt Warning: During execution of `Reduced\underline{ }SnoptAlgorithm::Solve`\\ the following Problem occurred!\\
To use this algorithm you need to have SNOPT installed! \\To use this set the WITH\underline{ }SNOPT CompilerFlag.}

### 2013-12-10 Examples/OPT/StatPDE/Example9/content.tex
Update Doku. Habe Winnis-Superfunktionen fuer den Moment auskommentiert, da ich trotz tex-Update sonst nicht kompilieren kann

\subsubsection{Program description}
\twick{
This example is similar to Example \ref{OPT_Stat_Param_Nonlin_Fluid} (both 
based on Example \ref{OPT_Stat_Param_Lin_Ellipt}), except that we 
have again much more complicated (nonlinear) equations. A modification 
of this example and several numerical tests are presented 
in \cite{RiWi13_fsi_opt}.
}

>>>>

\subsubsection{Program description}
This example is similar to Example \ref{OPT_Stat_Param_Nonlin_Fluid} (both 
based on Example \ref{OPT_Stat_Param_Lin_Ellipt}), except that we 
have again much more complicated (nonlinear) equations. A modification 
of this example and several numerical tests are presented 
in \cite{RiWi13_fsi_opt}.
